---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2025_spring")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```

```{r}

# print will take a string and return it in the terminal.
print("Hello World.")

```

```{r}

# This is a comment inside of a codeblock. Comments doesn't run when we run the codeblock.
# Adding this command is a good way to start your class reference notebook.

```

## read/load data - command changes slightly based on file type
```{r}

NAME(no hyphens) <- read_rds("data/NAME.rds")

```

# "data/NAME.rds" = file path

```{r}

NAME(no hyphens) <- read_csv("data/NAME.csv")

```

#preland 2
```{r}

#lubridate: a system that makes it to R Studio can read, extract, and work with dates and time data better.

# head(DATASETNAME): simplifies the presentation of the dataset.

# summary(DATASETNAME): provides a table summarizing the overall dataset.

# colnames(DATASETNAME):displays the column names.

#glimpse(DATASETNAME): provides an overall glimpse of the data (columns, rows and what data exists in those categories.)


```


# n(), the counting 
```{r}

n()

```


#better ordering
```{r}

#creates a new column/variable with `mutate` by using preexisting data, showing the percentage of votes made per candidate on election day. Then arranges the data by the percent_election_day variable.

primary_24 |>
  select(office_name, office_district, candidate_name, party, county_name, election_day, votes) |>
  mutate(
  percent_election_day = (election_day/votes)*100
)  |> 
  arrange(desc(percent_election_day))

```


#prelab 3 - mutates the <chr> data into <dbl> so that it is read as numbers, not as characters
```{r}

```{r}
texas_precinct_20 <- texas_precinct_20 |>
  mutate(election_day = case_when(
    election_day == '5+' ~ '49',
    .default = election_day
  ))

texas_precinct_20 <- texas_precinct_20 |>
  mutate(election_day = as.numeric(election_day))

glimpse(texas_precinct_20)
```

#renames the specified column
```{r}
DATASET |>
  rename(linenumber = x1_linenumber)

```

#find duplicate rows
```{r}

cleaned_conowingo |> 
  get_dupes()

```

#remove duplicate rows
```{r}

cleaned_conowingo |>


```

#cleans names, renames a column, transforms "amount" into the <dbl> data type, looks for distinct instances, mutates zip codes into their first 5 digits, and mutates all variables in the "city" column to display in title case.

```{r}
# cleaning functions
cleaned_DATASET <- DATASET |>
  clean_names() |> 
  rename(linenumber = x1_linenumber) |> 
  mutate(amount = as.numeric(amount)) |> 
  distinct() |>
  mutate(zip = str_sub(zip, start=1L, end=5L)) |>
  mutate(city = str_to_title(city))
  

# display the cleaned dataset
cleaned_conowingo

```

##casewhen: assign a column to multiple rows based on variables. basically makes a list (c) of which preexisting variables to assign a new column's variable to
```{r}
DATASET |>
  mutate(
    NEWCOLUMN = case_when(
      REFERENCEDCOLUMN %in% c("Variable1", "Var2", "Var 3") ~ "New Assigned Column"
    )
  )
```


##rename one variable without nuking the rest of them
```{r}

DATASET |>
  mutate(
    NEWCOLUMNNAME = case_when(
      COLUMNNAME == "variable" ~ "newVariableName",
      TRUE ~ COLUMNNAME
      ##^ "TRUE ~ COLUMNNAME" is what keeps the other variables in the column from getting nuked
      )
    ) 
```


##make all column names lowercase w/ underscores instead of spaces using library janitor
```{r}

library(janitor)

md_voters_2024 |>
  clean_names()

#etc etc etc. you can do this right when you import data

```

##remove junk columns
```{r}

select(-c(COL1, COL2, COL3, ETC))

```

#combine different data sets
```{r}
# bind_rows with list
NEW_DATASET_NAME <- bind_rows(list(DS1, DS2, DS3, DS4, DS5))
View(NEW_DATASET_NAME)
```
